---
title: "A4: Assessing Inequalities in United States Incarceration Data"
author: "Madeleine Clarke"
date: '2022-11-21'
output: html_document
---

```{r echo=FALSE, eval=TRUE, message=FALSE}
#source("../source/a4-helpers.R")
source("../source/analysis.R")
```

## Abstract 
This website explores inequalities in incarceration patterns within the United States. Analyses have been conducted using incarceration data collected by the Vera Institute, an advocacy and research group focused on transforming American criminal legal and immigration systems. The Vera Institute is particularly interested in ending mass incarceration and reforming our systems to be more just and fair for people of color and other marginalized groups. As such, the follow report will focus on revealing patterns of racial inequality and over-incarceration within the United States. 

## Section 1: Introduction 
The aim of this data project is to reveal injustices within United States incarceration trends. To accomplish this goal, I will utilize data from the Vera Institute that contains demographic information and jail population data at the county-level for all 50 states. More specifically, I aim to reveal patterns of racial inequality in incarceration rates, as well as highlight differences in incarceration patterns between states. 

The United States is home to the largest incarcerated population in the world. The over-policing and incarceration of people of color has been [well documented](https://www.pewresearch.org/fact-tank/2020/06/03/10-things-we-know-about-race-and-policing-in-the-u-s/), and has become a major topic in recent years as part of the resurgence of the Black Lives Matter movement. As it stands, the U.S. criminal justice system targets already vulnerable individuals and communities, and works to [perpetuate poverty and racial disparites](https://www.vera.org/ending-mass-incarceration/criminalization-racial-disparities) in wealth and health. Addressing inequalities within the criminal legal system in the United States is necessary to 

I want to acknowledge that the real-world data I am using for my analysis represents the lives of hundreds of thousands of individuals and their communities who have been affected by incarceration. I want to honor the hardships and emotions that were involved in becoming a data point in the dataset that I am utilizing. As such, I will ground my work in the values described below, as well as acknowledging those (both direct and indirect) who may be implicated in my analysis. 

* **Stakeholders:** 
  - Direct: families and communities that are affected by mass-incarceration, police systems, criminal and legal systems, jails and prisons.
  - Indirect: larger communities/social groups (e.g., people living in poverty, communities of color), individuals who will/are likely to interact with the criminal legal system in the future.
* **Values:** Throughout my work for this project I will engage with values of Justice, Liberation, Equity and Equality.


## Section 2: Data Summary 
[TODO]

## Section 3: Growth of the Prison Population
``` {r pop growth, include = FALSE}
source("../source/analysis.R")
```

``` {r plotting bar chart, echo = FALSE}
plot(jail_pop_graph)
```

This graph shows the growth of the total population of all jails in the United States between the years 1970 and 2018. One can note that the jail population was quite stable through the 1970's, hovering around 150,000 individuals. In the early 1980s, the U.S. jail population begins a steap and steady incline. The population continues to grow into the  2000s, reaching its peak of almost 800,000 individuals just before 2010. Since then, the U.S. jail population has decreased slightly, with the most recent counts standing at around 750,000 individuals.

## Section 4: Growth of Prison Population by State
``` {r state growth, include = FALSE}
source("../source/analysis.R")
```

``` {r plotting state growth, echo = FALSE}
plot(jail_pop_state)
```

This graph shows the growth of jail populations by state. This allows viewers to detect differences in the size of jail populations at the state-level. While the jail population of a particular state may be strongly related to overall population size, comparing states with similar overall population sizes helps to reveal differential patterns in incarceration rates. I've chosen to display both New York and Florida on this graph, which have very similar population sizes. However, Florida's jail population is almost double that of New York. On the other hand, two smaller states of almost equal size: New Hampshire and Maine have almost identical jail population sizes and growth trends. Washington and Michigan have high-to-mid range population sizes, and their jail populations appear to be roughly proportional.

## Section 5: Comparing the Representation of Black and White Individuals in Jail vs. the General Population
``` {r scatterplot, include = FALSE}
source("../source/analysis.R")
```

# ``` {r plot scatterplot, echo = FALSE}
# plot(bw_percentage_plot)
# ```

The above plot seeks to answer the question: *Are certain racial groups over-represented in jail populations compared to others?* To answer this question, I have created a scatterplot that compares the representation of Black and White people in jail versus their representation in the general population. Each dot represents population data from a singular U.S. county in 2018. For simplicity, only Black and White populations are displayed. The BLUE colored dots represent Black population percentages and the RED dots represent White population percentages.

In theory, for any given county the representation of a certain racial group in the general public should align with the representation of that racial group in jail. For example, if Black people make up ~7% of the general population in King County, WA then theoretically they should make up ~7% of the incarcerated population in King County. This means in a world with no racial inequalities in incarceration rates, the two colors of dots should generally cluster in the same way, and should be roughly along the x=y slope (see the diagonal line displayed on the graph). However, this is not what the data reveals. In fact, the data shows that Black individuals tend to cluster _above_ the diagonal line, and White individuals tend to cluster _below_ it. This means that on average, Black individuals are **over-represented** in jail populations across the U.S. compared to Whites.

**Note**: the breakdown of general population data by race was only available for ages 15-64. While this is a limitation of the dataset, I believe that

<!-- ## Section 6: <a map shows potential patterns of inequality that vary geographically> -->
<!-- ``` {r map, include = FALSE} -->
<!-- source("../source/analysis.R") -->
<!-- ``` -->

<!-- This [source](https://redbird.shinyapps.io/police-bias-map/) helped me with my conceptual model -->


_All the very best with your work!_

